\documentclass[11pt]{article}

% Change "review" to "final" to generate the final (sometimes called camera-ready) version.
% Change to "preprint" to generate a non-anonymous version with page numbers.
% \usepackage[review]{acl}
\usepackage[preprint]{acl}

% Standard package includes
\usepackage{times}
\usepackage{latexsym}

% For proper rendering and hyphenation of words containing Latin characters (including in bib files)
\usepackage[T1]{fontenc}
% For Vietnamese characters
% \usepackage[T5]{fontenc}
% See https://www.latex-project.org/help/documentation/encguide.pdf for other character sets

% This assumes your files are encoded as UTF8
\usepackage[utf8]{inputenc}

% This is not strictly necessary, and may be commented out,
% but it will improve the layout of the manuscript,
% and will typically save some space.
\usepackage{microtype}

% This is also not strictly necessary, and may be commented out.
% However, it will improve the aesthetics of text in
% the typewriter font.
\usepackage{inconsolata}

%Including images in your LaTeX document requires adding
%additional package(s)
\usepackage{graphicx}

% If the title and author information does not fit in the area allocated, uncomment the following
%
%\setlength\titlebox{<dim>}
%
% and set <dim> to something 5cm or larger.

\title{Handwritten Mathematical Expression Recognition Using Convolutional Neural Networks}

% Author information can be set in various styles:
% For several authors from the same institution:
% \author{Author 1 \and ... \and Author n \\
%         Address line \\ ... \\ Address line}
% if the names do not fit well on one line use
%         Author 1 \\ {\bf Author 2} \\ ... \\ {\bf Author n} \\
% For authors from different institutions:
% \author{Author 1 \\ Address line \\  ... \\ Address line
%         \And  ... \And
%         Author n \\ Address line \\ ... \\ Address line}
% To start a separate ``row'' of authors use \AND, as in
% \author{Author 1 \\ Address line \\  ... \\ Address line
%         \AND
%         Author 2 \\ Address line \\ ... \\ Address line \And
%         Author 3 \\ Address line \\ ... \\ Address line}

% Bios of the researchers! Include your name, NetID, department, and year in the
% program. Also list what each team member will likely be responsible for or how you
% plan to divide the work fairly.

\author{Adnan Binmahfoudh \\
    Computer Science \\
    Senior \\
    \\\And
    Daniel Conti \\
    Computer Science \\
    Senior \\
    \\\And
    Mihir Patel \\
    Data Science \\
    Junior \\
    \\\And
    Nathan Emeott \\
    Computer Science \\
    Junior \\
    \\\And
    Patrick Schrems \\
    Computational \& \\Applied Mathematics \\
    and Computer Science \\
    Junior \\
}

%\author{
%  \textbf{First Author\textsuperscript{1}},
%  \textbf{Second Author\textsuperscript{1,2}},
%  \textbf{Third T. Author\textsuperscript{1}},
%  \textbf{Fourth Author\textsuperscript{1}},
%\\
%  \textbf{Fifth Author\textsuperscript{1,2}},
%  \textbf{Sixth Author\textsuperscript{1}},
%  \textbf{Seventh Author\textsuperscript{1}},
%  \textbf{Eighth Author \textsuperscript{1,2,3,4}},
%\\
%  \textbf{Ninth Author\textsuperscript{1}},
%  \textbf{Tenth Author\textsuperscript{1}},
%  \textbf{Eleventh E. Author\textsuperscript{1,2,3,4,5}},
%  \textbf{Twelfth Author\textsuperscript{1}},
%\\
%  \textbf{Thirteenth Author\textsuperscript{3}},
%  \textbf{Fourteenth F. Author\textsuperscript{2,4}},
%  \textbf{Fifteenth Author\textsuperscript{1}},
%  \textbf{Sixteenth Author\textsuperscript{1}},
%\\
%  \textbf{Seventeenth S. Author\textsuperscript{4,5}},
%  \textbf{Eighteenth Author\textsuperscript{3,4}},
%  \textbf{Nineteenth N. Author\textsuperscript{2,5}},
%  \textbf{Twentieth Author\textsuperscript{1}}
%\\
%\\
%  \textsuperscript{1}Affiliation 1,
%  \textsuperscript{2}Affiliation 2,
%  \textsuperscript{3}Affiliation 3,
%  \textsuperscript{4}Affiliation 4,
%  \textsuperscript{5}Affiliation 5
%\\
%  \small{
%    \textbf{Correspondence:} \href{mailto:email@domain}{email@domain}
%  }
%}

\begin{document}
\maketitle

\section*{Author Bios and Responsibilities}

\begin{description}
	\item[\textbf{Adnan Binmahfoudh}] Senior in Computer Science. He will focus on dataset preparation and preprocessing.
	\item[\textbf{Daniel Conti}] Senior in Computer Science. He will handle the encoder implementation.
	\item[\textbf{Mihir Patel}] Junior in Data Science. He will be responsible for the Markdown decoder.
	\item[\textbf{Nathan Emeott}] Junior in Computer Science. He will work on the LaTeX decoder.
	\item[\textbf{Patrick Schrems}] Junior in Computational \& Applied Mathematics and Computer Science.
	      He will focus on data segmentation and SVM baseline models.
\end{description}

% % Force first indent
% \hspace*{\parindent}
% Adnan Binmahfoudh - Department of Computer Science (Senior)\\
% Responsible for LaTeX decoder.

% Daniel Conti - Department of Computer Science (Senior)\\
% Responsible for encoder.

% Mihir Patel - Department of Data Science (Junior)\\
% Responsible for dataset loading and preprocessing.

% Nathan Emeott - Department of Computer Science (Junior)\\
% Responsible for Markdown decoder.

% Patrick Schrems - Departments of Applied Mathematics \& Computer Science (Junior)\\
% Responsible for data segmentation and SVM baseline models.

\begin{abstract}
	Optical Character Recognition has seen many advancements recently, including improved accuracy via Large Language Models (LLMs),
	and enhanced layout understanding through computer vision techniques like structured segmentation. While these improvements have improved
	accuracy and complex format parsing, handwritten content still remains a challenge. Current methods for converting documents to
	Markdown and LaTeX mainly focus on digitally created or printed text, with little attention given to handwritten material.
	To combat this issue, we propose a multi-model approach, consisting of one Convolutional Neural Network encoder and two
	Recurrent Neural Network decoders: LaTeX and Markdown. First we segment the image, identifying text and math blocks.
	Then we choose the appropriate model according to the classification criteria; if the confidence does not meet a specific threshold,
	we use both models and select the best output. Further analyses are in progress. We expect this project to contribute to the improvement of
	handwritten content extraction.

	% …Traditional OCR relies heavily on image processing and pre-trained models, limiting performance with complex document layouts.
	% Deep learning models are able to adapt to uncommon document structures, unlike the pre-trained models commonly used by traditional OCR.
	% However traditional OCR is more lightweight and faster than the slower, more intense LLM-powered OCR….
\end{abstract}

\section{Introduction}

The problem we aim to solve with this project is converting handwritten text and math to Markdown and LaTeX.
Currently, many models can convert handwritten math into LaTeX, but the problem here is that many people write their calculations on paper.
The latest models do not usually support handwritten content or are largely inaccurate for such writing. This is where our underlying problem lies.
We aim to create a hybrid system that can deal with handwritten math and text and convert it into Markdown and LaTeX.
Primary users of our system would be students and instructors, or others who frequently write notes and calculations by hand.
The majority of students use paper for proofs and calculations, and converting them to LaTeX can be challenging, especially for new users.
Conversion can be time-consuming, as LaTeX code for equations can be quite long and complex. This is where our math-text hybrid conversion
approach would come into use, converting handwritten notes. Our solution removes the manual conversion time and skill challenge.
We propose a hybrid solution that classifies text and math, and then uses corresponding decoder models to extract the content.

\section{Related Works}

The task of converting handwritten mathematics into LaTeX has a few notable recent attempts. Both a Support Vector Machine (SVM) classifier and
Convolution Neural Network (CNN) showed success in converting individual characters to LaTeX \citep{chang2016painfree}. For handwriting samples with
multiple characters, SVM can also lead to misclassification in times where handwriting is highly diverse, such as the multiple ways of writing
the character `x,' in which a CNN performs significantly better \citep{schechter2017converting}.
Other testing indicates that encoder-decoder paradigms show promising results as they vastly improve on the ``counting problem,''
where complicated fractions or integrals lead to omitted or hallucinated characters as a result of the inherent non-linearity of mathematical expressions \citep{wang2020recognizinghandwrittenmathematicalexpressions}.
% TODO: Introduce Recurrent Neural Network (RNN) better
As transformers have become the dominant encoder-decoder technique, analysis comparing them to CNN-RNN
(Convolutional Neural Network--Recurrent Neural Network) methods shows potentially even better results \citep{sundararaj2024automatedlatexcodegeneration}.

While previous attempts use the single-encoder single-decoder paradigm, we intend to take it a step further by using two decoders.
This will allow us to do a full conversion of handwriting to Markdown and LaTeX as opposed to previous attempts solely converting to LaTeX.
We also noticed that ensemble modeling has been successful for optical character recognition \citep{preiss2025evaluationensemblelearningtechniques}.
Combining this with the diversity problem of handwritten mathematical expressions, we intend to explore techniques to boost model diversity in
ensemble approaches such as clustering as a preprocessing step for ensemble modeling, which has been successful in other
classification applications \citep{cui2021386}. Specifically, clustering the dataset and performing SVM on each resultant cluster,
only requiring each model to be responsible for a significantly smaller portion of the dataset overall.

\section{Dataset and Methodology}

% TODO: Reword
For our implementation, we will use a classifier to determine which output is more appropriate, with a confidence metric for each option.
If neither option reaches the minimum confidence metric, then both will be produced and the best output will be selected.
For handwritten math to LaTeX we use the MathWritingHandwritten dataset \citep{yo2025mathwritinghandwritten},
which has over 250k rows of handwritten mathematical expressions with the corresponding LaTeX code.
For parsing regular handwritten text we use the IAM Handwriting Database \citep{marti2002iam, teklia2024handwrittentext},
which has a large dataset of labeled handwritten text.

In our implementation we would use an Optical Character Recognition model, which uses a RNN and techniques such as
segmentation and contrast manipulation to isolate text. Then a binary classifier with a confidence metric could be used to determine
whether to produce LaTeX or Markdown. If neither option reaches the required threshold, both would be produced using the
respective decoders, and the best outcome would be used.
% TODO: Reword

\section{Classical Methods}

\subsection{Dataset and Preprocessing}

Our dataset and preprocessing pipeline that we have set up loads out two hugging face datasets. These datasets include MathWritingHandwritten for the math expressions, while the other dataset contains handwritten text. Our pipeline here works for both datasets, where we convert each image in our dataset to grayscale. We then resize it to be 28x28 pixels size, and then binarize our images using Otsu’s thresholding method. It is important to note that the current pipeline we have implemented works and was designed for SVM training right now. For our full pipeline, we went ahead and applied Gaussian blur for noise reduction in our images and then utilized adaptive thresholding for more cleaner binarization. It is important to note that our current implementation is the SVM pipeline only, which does not apply Gaussian blur. In our load\_dataset function, we use the svm\_loadimage function which follows the same process as our preprocess method in the preprocessing.py file. This function returns a list that follows: [text label, [binary pixels]]. For the binarization in this SVM function, we still use Otsu’s thresholding. Our main load\_dataset function uses the svm\_loadimage to get the label text and the binary pixels array for each image. We then build a new list for each image. Using the svm\_load\_image, we pull our pixels for each image. From here, the final list for each image that gets returned is: [[type label, text string], pixels]. The type label contains if it is either math for handwriting, and the text string is the actual text that is written, either latex math or handwriting.

\subsection{Segmentation}

To separate individual symbols from the handwritten image, we use a simple segmentation pipeline based on connected component analysis. First, the image is converted to grayscale and then binarized using Otsu’s thresholding, which helps distinguish the handwriting from the background. After this step, connected components are detected, where each group of connected pixels is treated as a potential symbol.
For each detected component, we compute a bounding box that captures the position and size of the symbol in the image. Very small components are filtered out to remove noise. The remaining bounding boxes are then sorted based on their position in the image so that symbols are processed in approximately left-to-right, top-to-bottom reading order. These segmented symbol regions are then passed to the next stage of the pipeline for feature extraction and classification using the SVM model.


\subsection{Support Vector Models}

For the SVM method, our pipeline is to have images loaded and pre-processed, segmented into individual symbols, classified by SVM models into symbol type (i.e., text vs LaTeX) and symbol, and displayed via a syntax tree. In our initial design, we have the SVM directly using pixels as features, but we intend to change this to well experimented features like Histogram of Oriented Gradients (HOG) in other image classification domains \citep{dalal1467360}. We also are using Principal Component Analysis to reduce the dimensionality of the images, particularly when areas like the corners likely hold very little information about the text. For clustering, we are using KMeans to cluster the data and then training individual SVM models on each cluster. This means that the final output in the ensemble architecture would be first classifying the unlabeled cluster of the image, loading that model, and predicting the output from that model. We also intend to try another ensemble architecture where first the Markdown/LaTeX class is predicted and then passed to a second SVM afterwards. For testing the models, we are reporting Accuracy, Precision, F1-Score, and Recall.

\subsection{Syntax Parsing and Output}

The last step of the process involves combining the predicted markdown and LaTeX together. This is done with an Abstract Syntax Tree (AST), which uses the information provided by segmentation and classification to produce the final output. Classical OCR methods commonly use this since they can't handle the multi-modal context that transformers are able to. However, to build the AST, we first need to find the formula line using the bounding box information from the classification stage. The baseline is a horizontal line through the symbols that dominate the equation. For example, the baseline for the equation $y=ax_{1}+bx_{2}^{2}$ would include they symbols $y=ax+bx$, since those are the base symbols of the formula. Superscripts and subscripts are treated as less dominant than the character that owns them. We take the symbol that dominates the others the most and construct the baseline from it. Then we build up the mathematical structural relationship between the others, expanding outwards from the baseline. Once we have the structural relationship between the symbols, we can build the final LaTeX output and add it to the final markdown text.


\section{Initial Experimental Results}

Despite the limitations we are now working around (see Limitations section), we ran a test of the two SVM architectures using MNIST digits and MNIST fashion with digits having labels [0, label] and fashion having labels [1, label]. With the total of 140,000 images, we split them into 80/20 training/test and got the following results:

\begin{table}[h!]
	\centering
	\begin{tabular}{|c|c|c|}
		\hline
		\textbf{Digit/Fashion} & \textbf{General SVM} & \textbf{Ensemble} \\
		\hline
		\verb|Accuracy|        & 0.9972               & 0.9981            \\\hline
		\verb|Precision|       & 0.9973               & 0.9981            \\\hline
		\verb|F1-Score|        & 0.9972               & 0.9981            \\\hline
		\verb|Recall|          & 0.9972               & 0.9981            \\\hline
	\end{tabular}

	\begin{tabular}{|c|c|c|}
		\hline
		\textbf{Label}   & \textbf{General SVM} & \textbf{Ensemble} \\
		\hline
		\verb|Accuracy|  & 0.7876               & 0.8435            \\\hline
		\verb|Precision| & 0.7870               & 0.8424            \\\hline
		\verb|F1-Score|  & 0.7849               & 0.8426            \\\hline
		\verb|Recall|    & 0.7876               & 0.8435            \\\hline
	\end{tabular}	\caption{Intial experimental results on the Digit/Fashion and Label methods respectively.}
\end{table}

Notably, while neither struggled under any metric with determining digit vs fashion, the ensemble SVM architecture outperformed the general SVM in all of the metrics with statistical significance (all p-values < 0.000001), indicating promising results for when we finally can switch to Markdown + LaTeX.

\section{Limitations}

As of now, our biggest limitation is the problem surrounding training the SVM on characters. For our pipeline when actually using the model, we have a structure in place for segmenting the images into characters, but this does not help us in training because our pipeline has no way of accurately separating the labels, especially for LaTeX. As of now, we are deciding on one of three options: train on labels our code splits with understanding that label accuracy will drop, train only on images that are labeled as a single symbol (e.g., a, A, $\alpha$, etc) with understanding that our dataset is now severely limited, or use other datasets to train the models and just use our current two as test datasets for comparison to the encoder-decoder architecture. As of the time of this short paper, we are going to use two new datasets, EMNIST for Markdown text \citep{cohen2017emnistextension} and HASYv2 for LaTeX \citep{thoma2017hasyv2}, and keep the alternatives as a backup plan in case everything fails on us.

Another limitation is that we have to manually try different cluster numbers for KMeans, and due to training times which we are already using PCA for, we are limited in the amount of clusters we can use. The idea is that we want clusters because they can make the data more separable, but there is also an upper limit because eventually the unsupervised KMeans model has a larger impact on the final output than each of the SVM models does. As of now, we are just planning on testing different cluster numbers until we get the one that performs best, starting with small amounts > 1 (e.g., 2, 3, 4, or 5).

\section{Planned Neural Network Method}

We are still planning on using an encoder-decoder architecture. Because of the snags we were running into, we diverted our focus towards getting the classical approach to work, and now that we are wrapping that up, we will start training the neural network method immediately. Because it can handle strings of text much easier, we expect to spend significantly less time preparing data for it with the trade off of more training time.

% \section*{Acknowledgments}

% Bibliography entries for the entire Anthology, followed by custom entries
%\bibliography{custom,anthology-overleaf-1,anthology-overleaf-2}

% Custom bibliography entries only
\bibliography{custom}

\end{document}
